---
title: "DAS Project 2 - Group 13"
format:
  html:
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This project analyses data from the Coffee Quality Database (CQD), which contains information on coffee batches and their quality-related characteristics. The aim is to investigate which features of coffee, such as aroma, flavor, acidity, defects and altitude, influence whether a batch is classified as Good or Poor.

To answer this question, the dataset is first cleaned and explored using descriptive summaries and visualisations. A generalised linear model (GLM) is then used to examine how these variables are associated with coffee quality classification.

# Data Cleaning

```{r}
library(tidyverse)

coffee <- read_csv("data/dataset13.csv", show_col_types = FALSE)
```

```{r}
glimpse(coffee)
summary(coffee)
colSums(is.na(coffee))
sum(duplicated(coffee))
```

```{r}
coffee <- coffee %>%
  distinct() %>%
  mutate(
    Qualityclass = as.factor(Qualityclass),
    country_of_origin = as.factor(country_of_origin)
  )
```

The dataset was inspected using `glimpse()` and `summary()` to understand the structure and distributions of the variables. Missing values were checked using `colSums(is.na())`, and duplicate rows were identified using `duplicated()`. Duplicate rows were removed to improve data quality before analysis.

```{r}
coffee_clean <- coffee %>%
  filter(!is.na(country_of_origin),
         !is.na(aroma),
         !is.na(flavor),
         !is.na(acidity),
         !is.na(category_two_defects),
         !is.na(altitude_mean_meters),
         !is.na(Qualityclass)) %>%
#Handling Outliers
  filter(altitude_mean_meters > 0,
         altitude_mean_meters < 3000,
         aroma > 0,   
         flavor > 0,  
         acidity > 0) %>% 
#Altitude Binning & Log Transformation
  mutate(
    altitude_category = case_when(
      altitude_mean_meters < 1250 ~ "Low",
      altitude_mean_meters >= 1250 & altitude_mean_meters <= 1500 ~ "Medium",
      altitude_mean_meters > 1500 ~ "High"
    ),
    altitude_category = factor(altitude_category, levels = c("Low", "Medium", "High")),
    log_defects = log(category_two_defects + 1)
  )
```

The dataset was refined by removing missing values and unrealistic outliers, applying a 3000m altitude threshold based on biological limits. The harvested variable and country_of_origin were excluded to eliminate climatic and geographical noise. Importantly, altitude was categorized into Low (\<1250m), Medium (1250-1500m), and High (\>1500m) based on domain literature. To facilitate modeling, altitude was categorized into three bins based on domain literature, and category_two_defects was log-transformed to stabilize its variance. This resulted in a robust and high-quality dataset ready for analysis.

```{r}
nrow(coffee)
nrow(coffee_clean)
```

The original dataset had 1,145 rows. We removed 217 rows that contained duplicates, missing values, or obvious errors (such as unrealistic altitude). This leaves us with 927 high-quality rows, which is a very solid sample size for building our predictive model.

# Exploratory Data Analysis

## Quality Class Proportion

```{r}
quality_stats <- coffee_clean %>%
  count(Qualityclass) %>%
  mutate(Percentage = round(n / sum(n) * 100, 1))
print(quality_stats)
```

The summary shows an almost perfectly balanced dataset, with approximately 50.9% of the coffees classified as "Good" and 49.1% as "Poor." This balance ensures that our subsequent logistic regression model will not suffer from bias toward a majority class.

## Descriptive Statistics & Distribution Check

```{r}
library(gridExtra)
# View summary statistics for key predictors
summary(coffee_clean %>% select(aroma, flavor, acidity, log_defects))
# Plot histograms to visualize the distribution shapes and long tails
p1 <- ggplot(coffee_clean, aes(x = aroma)) + 
  geom_histogram(fill="lightblue", color="white", bins=30) + 
  theme_minimal() + labs(title="Aroma Distribution", x="Aroma", y="Count")
p2 <- ggplot(coffee_clean, aes(x = flavor)) + 
  geom_histogram(fill="lightgreen", color="white", bins=30) + 
  theme_minimal() + labs(title="Flavor Distribution", x="Flavor", y="Count")
p3 <- ggplot(coffee_clean, aes(x = acidity)) + 
  geom_histogram(fill="lightpink", color="white", bins=30) + 
  theme_minimal() + labs(title="Acidity Distribution", x="Acidity", y="Count")
p4 <- ggplot(coffee_clean, aes(x = log_defects)) + 
  geom_histogram(fill="orange", color="white", bins=30) + 
  theme_minimal() + labs(title="Log(Defects+1) Distribution", x="Log(Defects+1)", y="Count")
# Arrange all four plots in a 2x2 grid for better readability
grid.arrange(p1, p2, p3, p4, ncol=2, nrow=2)
```

The histograms reveal that the sensory attributes (aroma, flavor, and acidity) roughly follow a normal distribution centered around 7.5, with a few natural extreme low scores acting as the left tail. For the defects variable, we applied a log(x+1) transformation to address its severe right-skewness. While the resulting log_defects distribution still has many zero values due to the high frequency of zero-defect batches, the transformation effectively compresses the long tail of outliers, making the data more stable for our modeling.

## Altitude Category Distribution

To ensure the reliability of our subsequent statistical modeling, it is essential to verify that our newly engineered `altitude_category` variable has sufficient observations within each group.

```{r}
# Calculate the count and percentage for each altitude category
altitude_stats <- coffee_clean %>%
  count(altitude_category) %>%
  mutate(Percentage = round(n / sum(n) * 100, 1))
print(altitude_stats)
# Visualize the distribution of altitude categories
ggplot(coffee_clean, aes(x = altitude_category, fill = altitude_category)) +
  geom_bar() +
  scale_fill_brewer(palette = "Blues") +
  theme_minimal() +
  labs(title = "Distribution of Coffee Batches by Altitude",
       x = "Altitude Category",
       y = "Number of Batches") +
  theme(legend.position = "none")
```

The distribution table and bar chart show that our altitude_category is well-balanced across all three levels. Low altitude has the most observations (38.6%), followed by Medium (30.2%) and High (31.2%). This balanced distribution is crucial as it ensures our GLM model has enough data in each group to produce statistically significant results, rather than being biased by a single dominant category.

## Correlation & Multicollinearity Check

Before fitting our Generalized Linear Model (GLM), it is vital to check for potential multicollinearity among our continuous predictors. High correlation between independent variables can lead to unstable coefficient estimates. We visualize this using a correlation heatmap.

```{r}
library(corrplot)
# Calculate the correlation matrix for continuous variables
cor_matrix <- coffee_clean %>%
  select(aroma, flavor, acidity, category_two_defects) %>%
  cor()
# Generate a professional correlation heatmap
corrplot(cor_matrix, method = "color", type = "upper", addCoef.col = "black", 
         tl.col = "black", tl.srt = 45, 
         title = "Correlation Matrix of Numeric Predictors", mar = c(0,0,1,0))
```

The heatmap indicates strong positive correlations among the sensory attributes, particularly between flavor and acidity ($r \approx 0.75$), and aroma and flavor ($r \approx 0.73$). While expected in coffee cupping (coffees that smell good typically taste good), this high correlation serves as an early warning. It dictates that we must rigorously monitor Variance Inflation Factors (VIF) during the model diagnostic phase to ensure our model remains stable.

## Exploring X vs Y Relationships

Finally, we visually explore how our primary predictors relate to our target variable (Qualityclass) to confirm their predictive potential. We focus on our strongest continuous predictor (Flavor) and our newly engineered categorical predictor (Altitude Category).

```{r}
# Plot 1: Flavor distribution across Quality Classes
p_flavor <- ggplot(coffee_clean, aes(x = Qualityclass, y = flavor, fill = Qualityclass)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Good" = "#66c2a5", "Poor" = "#fc8d62")) +
  labs(title = "Flavor vs Quality Class", x = "Quality Class", y = "Flavor Score") +
  theme_minimal() +
  theme(legend.position = "none")
# Plot 2: Quality Class proportion stacked by Altitude Category
p_altitude <- ggplot(coffee_clean, aes(x = altitude_category, fill = Qualityclass)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Good" = "#66c2a5", "Poor" = "#fc8d62")) +
  labs(title = "Proportion of Good/Poor Coffee by Altitude",
       x = "Altitude Category", y = "Percentage") +
  theme_minimal()
p_flavor
p_altitude
```

The visualizations confirm strong predictive associations. The boxplot demonstrates a stark distinction in flavor scores: batches classified as "Good" consistently score higher than "Poor" batches. Furthermore, the stacked bar chart reveals a striking linear trend regarding the growing environment: higher altitudes yield a significantly larger proportion of "Good" coffee (over 70% in the High category). These insights lay a robust foundation for our logistic regression modeling.

# Modelling

# Conclusion
